Inference for multivariate peaks over threshold models

### 2021-11-06 abstract.tex
overleaf updates

{\bf We need an abstract}
Extreme analysis

Drawing on the constructive definition for the multivariate Pareto introduced in~\cite{ferreira2014} which separates the radial component of the RV from the angular, we introduce a method for inferring the distribution of the angular component.

>>>>

{\bf We need an abstract}

The constructive definition of the multivariate Pareto introduced 
    in~\cite{ferreira2014} separates the Pareto RV $\bm{z}\in {\mathbb R}^d\not[0,1]^d$
    into independent radial,  $r \in (1,\infty)$ and angular 
    $\bm{y} \in {\mathbb S}_{\infty}^{d-1}$ components.
    This paper proposes a method for inferring a density on ${\mathbb S}_{\infty}^{d-1}$
    as the limit of a series of manifolds, ${\mathbb S}_{p}^{d-1}$.  We project a
    distribution in ${\mathbb R}_+^d$ to ${\mathbb S}_p^{d-1}$; then as $p$ grows
    large, we approach a density on ${\mathbb S}_{\infty}^{d-1}$.

We discuss model selection criteria appropriate to ${\mathbb S}_{\infty}^{d-1}$.  We
    use the energy score criterion introduced in \cite{gneiting2007}, incorporating a
    novel kernel metric we develop suitable to the space.  We use this energy score, along
    with posterior predictive loss to inform model choice in how well we recover the distribution of simulated data.  We then fit these models to real data

%

### 2021-11-06 evaluation.tex
overleaf updates

In order to assess and compare the estimation of a distribution on ${\mathbb S}_\infty^{d-1}$ we consider the theory of proper scoring rules developed in \cite{gneiting2007}.
As mentioned in Section~\ref{subsubsec:norm}, our approach does not provide a density on ${\mathbb S}_{\infty}^{d-1}$, restricting our ability to construct model selection criteria to sample-based approaches.  In this manner, we consider two methods of scoring:  \emph{posterior predictive loss} (PPL) introduced in \cite{gelfand1998}, and \emph{energy score} (ES) introduced in \cite{gneiting2007}.

\subsection{Posterior predictive loss}
The posterior predictive loss criterion is introduced in \cite{gelfand1998}.  When we
  assume a squared error loss function, then for the $i$th observation, the posterior predictive
  loss criterion is computed as
  \begin{equation}
    \label{eq:ppl}
      S_{\text{PPL}}^k\left(P,x_i\right) = \text{Var}_P\left(X_i\right) + \frac{k}{k+1}\left(\text{E}_P\left[X_{i}\right] - x_i\right)^2
  \end{equation}
  where $X_i$ is a random variable from the posterior predictive distribution for $x_i$.  The
  scalar $k$ is a weighting factor by which we can scale the importance of goodness of fit
  relative to precision.  In our analysis, we take the limit as $k\to\infty$, weighting both
  equally.  Interpreting this criterion, a smaller $\text{Var}(X_{i})$ indicates a higher
  precision, and a smaller $(\text{E}[X_{i}] - x_{i})^2$  indicates a better fit.  Thus,
  smaller is better.  Note that this is defined for a univariate distribution.  As we are dealing
  with a multivariate distribution, we will be taking the posterior predictive loss
  summed over all dimensions.  That is,
  \begin{equation}
    \label{eq:ppl2}
      S_{\text{PPL}}\left(P,\bm{x}_i\right) = \frac{1}{d}\sum_{\ell = 1}^d\left[\text{Var}_P\left(X_{i\ell}\right) + \left(\text{E}_P\left[X_{i\ell}\right] - x_{i\ell}\right)^2\right]
  \end{equation}
  Then we report the average $S$, taken over all $i$.

@@

Evaluating $g$ as defined in Proposition \ref{prop:g} requires the solution of a $d-2$-dimensional 
  optimization problem in order to obtain $\bm{c}$.  However, a computationally easier \makenote{rewrite sentence} method exists.
\begin{prop}
    Let $\bm{a} \in {\mathbb C}_{\ell}^{d-1}$, and $\bm{b} \in {\mathbb C}_{\jmath}^{d-1}$.  The 
  rotation $P_{\jmath\ell}(\cdot)$ required to move the $\jmath$th face along the $\ell$th axis 
  produces the vector $\bm{b}^\prime$, where
  \begin{equation}
    \label{eqn:rotation}
    P_{\jmath\ell}(\bm{b}) = \bm{b}^{\prime}_i = 
    \begin{cases}
        b_{i} &\text{for }i\neq \jmath,\ell\\
        1 &\text{for }i = \ell\\
        2 - b_{\ell} &\text{for }i = \jmath
    \end{cases}.
  \end{equation}
  Then $g(\bm{a},\bm{b}) = \pnorm{\bm{a} - \bm{b}^{\prime}}{2}$.
\end{prop}
{\em Proof:}
Note that for 
  $\bm{c} \in {\mathbb C}_{\jmath\ell}^{d-2} = {\mathbb C}_{\jmath}^{d-1}\cap{\mathbb C}_{\ell}^{d-1}$,
  $P_{\jmath\ell}(\bm{c}) = \bm{c}$.  Then
  \begin{equation*}
     \pnorm{\bm{b} - \bm{c}}{2} = \pnorm{P_{\jmath\ell}(\bm{b} - \bm{c})}{2} = \pnorm{\bm{b}^{\prime} - \bm{c}}{2},
  \end{equation*}
  as rotations preserve Euclidean norms.  Then,
  \begin{equation*}
    \begin{aligned}
    g(\bm{a}, \bm{b}) &= \pnorm{\bm{a} - \bm{c}}{2} + \pnorm{\bm{b} - \bm{c}}{2}\\
    &= \pnorm{\bm{a} - \bm{c}}{2} + \pnorm{\bm{b}^{\prime} - \bm{c}}{2} = \pnorm{\bm{a} - \bm{b}^{\prime}}{2}
    \end{aligned}
  \end{equation*}
  when $\bm{c} \in \overrightarrow{\bm{a}\bm{b}^{\prime}}$, as the lower bound of the triangle inequality.  We have already declared $\bm{c}\in{\mathbb C}_{\jmath\ell}^{d-2}$, so it must reside at $\overrightarrow{\bm{a}\bm{b}^{\prime}}\cap{\mathbb C}_{\jmath\ell}^{d-2}$.  Because the straight line between $\bm{a}$ and $\bm{b}^{\prime}$ passes through this point without further analysis, the optimization of $\bm{c}$ is accomplished \emph{en~passant}.$\hfill\Box$

Interpretation of the energy score as we have formulated it here is the same as that of posterior predictive loss: smaller is better.

>>>>

In order to assess and compare the estimation of a distribution on ${\mathbb S}_\infty^{d-1}$ we consider the theory of proper scoring rules developed in \cite{gneiting2007}.
As mentioned in Section~\ref{subsec:projgamma}, our approach does not provide a density on ${\mathbb S}_{\infty}^{d-1}$, restricting our ability to construct model selection criteria to sample-based approaches.  We consider two scoring methods:  \emph{posterior predictive loss} (PPL) introduced in \cite{gelfand1998}, and \emph{energy score} (ES) introduced in \cite{gneiting2007}.

\subsection{Posterior predictive loss}
The posterior predictive loss criterion is introduced in \cite{gelfand1998}.  When we
  assume a squared error loss function, then for the $i$th observation, the posterior predictive
  loss criterion is computed as
  \begin{equation}
    \label{eq:ppl}
      S_{\text{PPL}}^k\left(P,x_i\right) = \text{Var}_P\left(X_i\right) + \frac{k}{k+1}\left(\text{E}_P\left[X_{i}\right] - x_i\right)^2
  \end{equation}
  where $X_i$ is a random variable from the posterior predictive distribution for $x_i$.  The
  scalar $k$ is a weighting factor that scales the importance of goodness of fit
  relative to precision.  In fact a small $\text{Var}(X_{i})$ indicates high
  precision, and a small $(\text{E}[X_{i}] - x_{i})^2$  indicates a good fit.  In our analysis, we take the limit as $k\to\infty$, weighting both
  equally. Overall, the smaller the score, the better.  Note that the PPLC score is defined for a univariate distribution.  As we are dealing
  with a multivariate distribution, we will be taking the posterior predictive loss
  summed over all dimensions.  That is,
  \begin{equation}
    \label{eq:ppl2}
      S_{\text{PPL}}\left(P,\bm{x}_i\right) = \frac{1}{d}\sum_{\ell = 1}^d\left[\text{Var}_P\left(X_{i\ell}\right) + \left(\text{E}_P\left[X_{i\ell}\right] - x_{i\ell}\right)^2\right] .
  \end{equation}

@@

Evaluating $g$ as defined in Proposition \ref{prop:g} requires the solution of a $d-2$-dimensional 
  optimization problem in order to obtain $\bm{c}$.  However, a computationally easier \makenote{rewrite sentence} method exists.
\begin{prop}
    Let $\bm{a} \in {\mathbb C}_{\ell}^{d-1}$, and $\bm{b} \in {\mathbb C}_{\jmath}^{d-1}$.  The 
  rotation $P_{\jmath\ell}(\cdot)$ required to rotate the $\jmath$th face along the $\ell$th axis 
  produces the vector $\bm{b}^\prime$, where
  \begin{equation}
    \label{eqn:rotation}
     \bm{b}^{\prime}_i = P_{\jmath\ell}(\bm{b}) = 
    \begin{cases}
        b_{i} &\text{for }i\neq \jmath,\ell\\
        1 &\text{for }i = \ell\\
        2 - b_{\ell} &\text{for }i = \jmath
    \end{cases}.
  \end{equation}
  Then $g(\bm{a},\bm{b}) = \pnorm{\bm{a} - \bm{b}^{\prime}}{2}$.
\end{prop}
{\em Proof:}
Note that for 
  $\bm{c} \in {\mathbb C}_{\jmath\ell}^{d-2} = {\mathbb C}_{\jmath}^{d-1}\cap{\mathbb C}_{\ell}^{d-1}$,
  $P_{\jmath\ell}(\bm{c}) = \bm{c}$.  Then
  \begin{equation*}
     \pnorm{\bm{b} - \bm{c}}{2} = \pnorm{P_{\jmath\ell}(\bm{b} - \bm{c})}{2} = \pnorm{\bm{b}^{\prime} - \bm{c}}{2},
  \end{equation*}
  as rotations preserve Euclidean norms.  Then,
  \begin{equation*}
    \begin{aligned}
    g(\bm{a}, \bm{b}) &= \pnorm{\bm{a} - \bm{c}}{2} + \pnorm{\bm{b} - \bm{c}}{2}\\
    &= \pnorm{\bm{a} - \bm{c}}{2} + \pnorm{\bm{b}^{\prime} - \bm{c}}{2} = \pnorm{\bm{a} - \bm{b}^{\prime}}{2}
    \end{aligned}
  \end{equation*}
  when $\bm{c} \in \overrightarrow{\bm{a}\bm{b}^{\prime}}$, as the lower bound of the triangle inequality.  We have already declared $\bm{c}\in{\mathbb C}_{\jmath\ell}^{d-2}$, so it must reside at $\overrightarrow{\bm{a}\bm{b}^{\prime}}\cap{\mathbb C}_{\jmath\ell}^{d-2}$.  Because the straight line between $\bm{a}$ and $\bm{b}^{\prime}$ passes through this point without further analysis, the optimization of $\bm{c}$ is accomplished \emph{en~passant}.$\hfill\Box$

We report the average Interpretation of the energy score as we have formulated it here is the same as that of posterior predictive loss: smaller is better.
